Type theory in type theory with single substitutions

### 2024-09-17 types_2024_postproc/p.tex
stuff

\newcommand{\ra}{\rightarrow}
\newcommand{\Ra}{\Rightarrow}
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\Tel}{\mathsf{Tel}}
\newcommand{\Tms}{\mathsf{Tms}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\ext}{\mathop{\triangleright}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\lam}{\mathsf{lam}}
\newcommand{\app}{\mathsf{app}}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\cd}{\mathsf{c}}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}} %from the book
\renewcommand{\tt}{\mathsf{tt}}
\newcommand{\fst}{\mathsf{fst}}
\newcommand{\snd}{\mathsf{snd}}
\newcommand{\Lift}{\mathsf{Lift}}
\newcommand{\mk}{\mathsf{mk}}
\newcommand{\un}{\mathsf{un}}

@@

@@

We add the rules for universes. A universe is a type containing codes
for types, this is witnessed by $\El$ (elements) and $\cd$ (code),
which make an isomorphism between terms of type $\U\,i$ and types of
level $i$. All three operations come with substitution rules ($\El[]$
and $\cd[]$ are interderivable, so it would be enough to assume one of
the two, but we add both to completely align with the algorithm
\cite{DBLP:conf/fscd/KaposiX24} generating the rules from the SOGAT
(\ref{eq:tt})).
\begin{alignat*}{10}
  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i)\hspace{2em} && \U[] && : (\U\,i)[\gamma] = \U\,i \\
  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i && \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) && \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma]) \\
  & \U\beta && : \El\,(\cd\,A) = A \\
  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A}
\end{alignat*}
Finally, we add the rules for moving types one level up. Because we
defined function space only between types at the same level, without
lifting, we cannot even define the polymorphic identity function.
\begin{alignat*}{10}
  & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i)\hspace{3.5em} && (\Lift\,A)[\gamma] = \Lift\,(A[\gamma]) \\
  & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) && (\mk\,a)[\gamma] = \mk\,(a[\gamma]) \\
  & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A && (\un\,a)[\gamma] = \un\,(a[\gamma]) \\
  & \Lift\beta && : \un\,(\mk\,a) = a \\
  & \Lift\eta && : \mk\,(\un\,a) = a
\end{alignat*}
This concludes the first-order single substitution semantics of the
SOGAT \ref{eq:tt}. For reference, we list the same rules again in
Appendix \ref{app:tt}. The theory can be extended with new type and
term formers in the same fashion: each rule has to be indexed over
arbitrary contexts and come equipped with a substitution rule. There
is no need to add more structural (substitution calculus) rules.

@@

Finally, we will remove equation $[\p^+][\langle\q\rangle]$ by making
$\Pi\eta$ conditional. This needs another change: we substitute
$\Pi\beta$ for a more general variant which is also conditional on the
same equation.
\begin{alignat*}{10}
  & \Pi\eta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra t =_e \lam\,(t[\p]\cdot\q) \\
  & \Pi\beta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra (\lam\,b)[\p]\cdot\q =_e b
\end{alignat*}
But first we need to check that the usage of $\Pi\beta$ above when
proving $B[\langle a\rangle][\gamma] = B[\gamma^+][\langle
  a[\gamma]\rangle]$ is derivable from $\Pi\beta'$. Assuming a
$\hat{B} : \Tm\,(\Gamma\ext A)\,\U$,
\begin{alignat*}{10}
  & \Pi\beta^\U :{} &&  \lam\,\hat{B}\cdot a && {=}(\q[\langle\rangle]) \\
  & && \lam\,\hat{B}\cdot(\q[\langle a\rangle])\,\, && {=}([\p][\langle\rangle]) \\
  & && ((\lam\,\hat{B})[\p][\langle a\rangle])\cdot(\q[\langle a\rangle])\,\, && {=}(\cdot[]'\,\U[]) \\
  & && ((\lam\,\hat{B})[\p]\cdot\q)[\langle a\rangle] && {=}(\Pi\beta'\,\U[]) \\
  & && \hat{B}[\langle a\rangle].
\end{alignat*}
Now we derive the assumption of $\Pi\beta/\eta'$ for any type $B : \Ty\,(\Gamma\ext A)$.
\begin{alignat*}{10}
  & B[\p^+][\langle\q\rangle] && {=}(\U\beta) \\
  & \El\,(\cd\,B)[\p^+][\langle\q\rangle] && {=}(\El[]) \\
  & \El\,(\cd\,B[\p^+][\langle\q\rangle]) && {=}(\Pi\beta^\U) \\
  & \El\,((\lam\,(\cd\,B[\p^+])\cdot\q) && {=}(\lam[]) \\
  & \El\,((\lam\,(\cd\,B)[\p]\cdot\q) && {=}(\Pi\beta'\,\U[]) \\
  & \El\,(\cd\,B) && {=}(\U\beta) \\
  & B
\end{alignat*}
This concludes the derivation of an even more minimalistic first-order
theory for the SOGAT (\ref{eq:tt}), relying essentially on the
universe and $\Pi$ types. It is clear from the construction that the
resulting theory is isomorphic to the one presented in Section
\ref{sec:tt} (and Appendix \ref{app:tt}), all equations are
interderivable. For reference, we list all the rules of the minimised
calculus in Appendix \ref{app:minimisation}.

TODO: summarise this section with a theorem?

@@

We can emulate the parallel substitution calculus using function
space, U and Sigma types

termification in the shallow paper is shown equationally complete

intuition, similar to minimisation: the subst calc is just there to
bootstrap U and Pi, those can do everything.

@@

@@

For reference, we list the rules of the theory introduced in Section
\ref{sec:tt}. The core substitution calculus:
\begin{alignat*}{10}
  & \Con && : \Set \\
  & \Ty && : \Con\ra\N\ra\Set \\
  & \diamond && : \Con \\
  & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con \\
  & \Sub && : \Con\ra\Con\ra\Set \\
  & \Tm && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set \\
  & \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
  & \langle\blank\rangle && : \Tm\,\Gamma\,A\ra\Sub\,\Gamma\,(\Gamma\ext A) \\
  & \blank^+ && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A) \\
  & \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \\
  & \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma]) \\
  & \q && : \Tm\,(\Gamma\ext A)\,(A[\p]) \\
  & [\p][^+] && : B[\p][\gamma^+] = B[\gamma][\p] \\
  & [\p][^+] && : b[\p][\gamma^+] = b[\gamma][\p] \\
  & \q[^+] && : \q[\gamma^+] = \q \\
  & [\p][\langle\rangle] && : B[\p][\langle a\rangle] = B \\
  & [\p][\langle\rangle] && : b[\p ][\langle a\rangle] = b \\
  & \q[\langle\rangle] && : \q[\langle a\rangle] = a \\
  & [\langle\rangle][] && : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle] \\
  & [\p^+][\langle\q\rangle] && : B[\p^+][\langle\q\rangle] = B
\end{alignat*}
Rules for individual type formers.
\begin{alignat*}{10}
  & \Pi && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
  & \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+]) \\
  & \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
  & \lam[] && : (\lam\,b)[\gamma] = \lam\,(b[\gamma^+]) \\
  & \blank\cdot\blank && : \Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,(B[\langle a\rangle]) \\
  & {\cdot}[] && : (t\cdot a)[\gamma] = (t[\gamma])\cdot(a[\gamma]) \\
  & \Pi\beta && : \lam\,b\cdot a = b[\langle a\rangle] \\
  & \Pi\eta && : t = \lam\,(t[\p]\cdot\q) \\
  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i) \\
  & \U[] && : (\U\,i)[\gamma] = \U\,i \\  
  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i \\
  & \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) \\
  & \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma]) \\
  & \U\beta && : \El\,(\cd\,A) = A \\
  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A} \\
  & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i) \\
  & \Lift[] && : (\Lift\,A)[\gamma] = \Lift\,(A[\gamma])
\end{alignat*}
\begin{alignat*}{10}
  & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) \\
  & \mk[] && : (\mk\,a)[\gamma] = \mk\,(a[\gamma]) \\
  & \un[] && : (\un\,a)[\gamma] = \un\,(a[\gamma]) \\
  & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A \\
  & \Lift\beta && : \un\,(\mk\,a) = a \\
  & \Lift\eta && : \mk\,(\un\,a) = a
\end{alignat*}

>>>>

\newcommand{\ra}{\rightarrow}
\newcommand{\Ra}{\Rightarrow}
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\Tel}{\mathsf{Tel}}
\newcommand{\Tms}{\mathsf{Tms}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\ext}{\mathop{\triangleright}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\lam}{\mathsf{lam}}
\newcommand{\app}{\mathsf{app}}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\cd}{\mathsf{c}}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}} %from the book
\renewcommand{\tt}{\mathsf{tt}}
\newcommand{\fst}{\mathsf{fst}}
\newcommand{\snd}{\mathsf{snd}}
\newcommand{\Lift}{\mathsf{Lift}}
\newcommand{\mk}{\mathsf{mk}}
\newcommand{\un}{\mathsf{un}}
\newcommand{\id}{\mathsf{id}}

@@

\newtheorem{problem}[theorem]{Problem}
\theoremstyle{remark}
\newtheorem{construction}[theorem]{Construction}

@@

We add the rules for universes. A universe is a type containing codes
for types, this is witnessed by $\El$ (elements) and $\cd$ (code),
which make an isomorphism between terms of type $\U\,i$ and types of
level $i$. All three operations come with substitution rules ($\El[]$
and $\cd[]$ are interderivable, so it would be enough to assume one of
the two, but we add both to completely align with the algorithm
\cite{DBLP:conf/fscd/KaposiX24} generating the rules from the SOGAT
(\ref{eq:tt})).
\begin{alignat*}{10}
  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i)\hspace{2em} && \U[] && : (\U\,i)[\gamma] = \U\,i \\
  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i && \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) && \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma]) \\
  & \U\beta && : \El\,(\cd\,A) = A \\
  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A}
\end{alignat*}
Finally, we add the rules for moving types one level up. Because we
defined function space only between types at the same level, without
lifting, we cannot even define the polymorphic identity function.
\begin{alignat*}{10}
  & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i)\hspace{3.5em} && (\Lift\,A)[\gamma] = \Lift\,(A[\gamma]) \\
  & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) && (\mk\,a)[\gamma] = \mk\,(a[\gamma]) \\
  & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A && (\un\,a)[\gamma] = \un\,(a[\gamma]) \\
  & \Lift\beta && : \un\,(\mk\,a) = a \\
  & \Lift\eta && : \mk\,(\un\,a) = a
\end{alignat*}
\begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$]\label{def:tt}
This concludes the first-order single substitution semantics of the
SOGAT (\ref{eq:tt}). For reference, we list the same rules again in
Appendix \ref{app:tt}.
\end{definition}
The theory can be extended with new type and term formers in the same
fashion: each rule has to be indexed over arbitrary contexts and come
equipped with a substitution rule. There is no need to add more
structural (substitution calculus) rules. For example, we show how to
extend Definition \ref{def:tt} with the SOGAT-rules (\ref{eq:sigma}).
\begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$, $\top$, and $\Sigma$ types]\label{def:sigma}
  We extend Definition \ref{def:tt} with the following rules.
  \begin{alignat*}{10}
    & \top && : \Ty\,\Gamma\,0 && \top[] && : \top[\gamma] = \top \\
    & \tt && : \Tm\,\Gamma\,\top && \tt[] && : \tt[\gamma] = \tt \\
    & \top\eta && : t = \tt \\
    & \Sigma && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
    & \Sigma[] && : (\Sigma\,A\,B)[\gamma] = \Sigma\,(A[\gamma])\,(B[\gamma^+]) \\
    & \fst,\snd && : \Tm\,\Gamma\,(\Sigma\,A\,B)\cong \\
    & && \rlap{$\hphantom{{}:{}} (a:\Tm\,\Gamma\,A)\times\Tm\,\Gamma\,(B[\langle a\rangle]) : (\blank,\blank)$} \\
    & {,}[] && : (a,b)[\gamma] = (a[\gamma],b[\gamma])
  \end{alignat*}
  To save space, we compressed the introduction, elimination, $\beta$
  and $\eta$ rules for $\Sigma$ into one isomorphism, and did not list
  substitution laws for $\fst$ and $\snd$ as they are derivable.
\end{definition}

@@

Finally, we will remove equation $[\p^+][\langle\q\rangle]$ by making
$\Pi\eta$ conditional. This needs another change: we substitute
$\Pi\beta$ for a more general variant which is also conditional on the
same equation.
\begin{alignat*}{10}
  & \Pi\eta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra t =_e \lam\,(t[\p]\cdot\q) \\
  & \Pi\beta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra (\lam\,b)[\p]\cdot\q =_e b
\end{alignat*}
But first we need to check that the usage of $\Pi\beta$ above when
proving $B[\langle a\rangle][\gamma] = B[\gamma^+][\langle
  a[\gamma]\rangle]$ is derivable from $\Pi\beta'$. Assuming a
$\hat{B} : \Tm\,(\Gamma\ext A)\,\U$,
\begin{alignat*}{10}
  & \Pi\beta^\U :{} &&  \lam\,\hat{B}\cdot a && {=}(\q[\langle\rangle]) \\
  & && \lam\,\hat{B}\cdot(\q[\langle a\rangle])\,\, && {=}([\p][\langle\rangle]) \\
  & && ((\lam\,\hat{B})[\p][\langle a\rangle])\cdot(\q[\langle a\rangle])\,\, && {=}(\cdot[]'\,\U[]) \\
  & && ((\lam\,\hat{B})[\p]\cdot\q)[\langle a\rangle] && {=}(\Pi\beta'\,\U[]) \\
  & && \hat{B}[\langle a\rangle].
\end{alignat*}
Now we derive the assumption of $\Pi\beta/\eta'$ for any type $B : \Ty\,(\Gamma\ext A)$.
\begin{alignat*}{10}
  & B[\p^+][\langle\q\rangle] && {=}(\U\beta) \\
  & \El\,(\cd\,B)[\p^+][\langle\q\rangle] && {=}(\El[]) \\
  & \El\,(\cd\,B[\p^+][\langle\q\rangle]) && {=}(\Pi\beta^\U) \\
  & \El\,((\lam\,(\cd\,B[\p^+])\cdot\q) && {=}(\lam[]) \\
  & \El\,((\lam\,(\cd\,B)[\p]\cdot\q) && {=}(\Pi\beta'\,\U[]) \\
  & \El\,(\cd\,B) && {=}(\U\beta) \\
  & B
\end{alignat*}
We summarise this section by formally stating what we defined.

\begin{definition}[Minimised single substitution calculus with $\Pi$ and $\U$]\label{def:min}
  This section defined the minimised first-order version of the SOGAT
  (\ref{eq:tt}), relying essentially on the presernce of universes and
  $\Pi$ types. For reference, list all the rules of the minimised
  calculus in Appendix \ref{app:minimisation}.
\end{definition}
\begin{theorem}
  The GATs of Definition \ref{def:min} and Definition \ref{def:tt} are
  isomorphic, in particular, all equations are interderivable.
\end{theorem}
\begin{proof}
  Clear from the construction in this section.
\end{proof}

@@

In this section we show that if the single substitution calculus has
certain type formers, then it is actually weakly equivalent to CwFs
with the same type formers. The idea is that
\begin{itemize}
\item using $\Sigma$ types we emulate contexts,
\item using functions between these $\Sigma$ types we emulate parallel
  substitutions,
\item using the functions into the universe we emulate dependent types,
\end{itemize}
and then single substitutions are just there to set up $\Sigma$, $\Pi$
and $\U$, and these types are enough to bootstrap the parallel
substitution calculus.

\begin{definition}[CwF]
  A category with $\N$-many families is defined as a category (objects
  denoted $\Con$, morphisms $\Sub$) with a terminal object $\diamond$
  ($\epsilon$ denotes the unique morphism into it); for each $i$ a
  presheaf of types (action on objects denoted $\Ty\,\blank\,i$,
  action on morphisms $\blank[\blank]$); for each $i$ a locally
  representable dependent presheaf $\Tm$ over $\Ty\,\blank\,i$
  (actions denoted $\Tm$, $\blank[\blank]$, local representability is
  denoted $\blank\ext\blank : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con$
  with an isomorphism
  $
  (\p\circ\blank,\q[\blank]) : \Sub\,\Delta\,(\Gamma\ext A) \cong (\gamma:\Sub\,\Delta\,\Gamma)\times\Tm\,\Delta\,(A[\gamma]) : (\blank,\blank)
  $
  natural in $\Delta$.
\end{definition}
\begin{problem}
  Every CwF (with type formers $\Pi, \dots, \Sigma$) is an SSC (with
  the same type formers, see Definition \ref{def:sigma}).
\end{problem}
\begin{proof}[Construction]
  Most operations are the same, we set $\gamma^+ :=
  (\gamma\circ\p,\q)$ and $\langle a\rangle := (\id,a)$.
\end{proof}
The following construction is also known as the standard model,
metacircular interpretation \cite{DBLP:conf/popl/AltenkirchK16},
contextualisation \cite{DBLP:conf/fscd/BocquetKS23}. Following
\cite{kaposi-shallow} we call it termification, as most sorts in the
new model are terms in the old model.
\begin{problem}[Termification]
  From a model of SSC (with $\Pi, \dots, \Sigma$, see Definition
  \ref{def:sigma}), we define a CwF (with the same type formers).
\end{problem}
\begin{proof}[Construction]
  We define iterated lifting $\Lift^k :
  \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(k+i)$ by induction on $k$, together
  with $\un^k : \Tm\,(\Lift^k\,A)\cong\Tm\,A : \mk^k$. The category
  part of the CwF is given by types in the empty context and functions
  between them, suitably lifted (on the left hand side of $:=$ there
  is the component in the new CwF-model, on the right hand side the
  components refer to the old SSC-model):
    & \Con && := (i:\N)\times\Ty\,\diamond\,i \\
    & \Sub\,\Delta\,\Gamma && := \Tm\,\diamond\,(\Lift^{\Gamma-\Delta}\,\Delta\Ra\Lift^{\Delta-\Gamma}\,\Gamma)
  In the definition of $\Sub$, we did not write the projections for
  $\Con$, so $\Delta$ can mean $\Delta_{.1}$ or $\Delta_{.2}$
  depending on the context, and we used the truncating subtraction of
  natural numbers. Composition of substitutions is quite involved, but
  it is just function composition written with explicit weakenings and
  appropriate (un)lifting. The category laws hold. The empty context is
  modelled by $\top$, its $\eta$ law holds 
    & \gamma\circ\delta && := \lam\Bigg(\mk^{\Theta-\Gamma}\,\bigg(\un^{\Delta-\Gamma}\,\Big(\gamma[\p]\cdot\mk^{\Gamma-\Delta}\,\big(\un^{\Theta-\Delta}\, \\
    & && \hspace{10em}(\delta[\p]\cdot\mk^{\Delta-\Theta}\,(\un^{\Gamma-\Theta}\,\q))\big)\Big)\bigg)\Bigg) \\
    & \rlap{$\id := \lam\,\q \hspace{4.35em} \diamond := (0,\top) \hspace{4.35em} \epsilon := \lam\,(\mk^\Gamma\,\tt)$}
  Types are given by functions into $\U$, terms are dependent
  functions into the type, with lots of lifting adjustments.
    & \Ty\,\Gamma\,i && := \Tm\,\diamond\,(\Lift^{1+i-\Gamma}\,\Gamma\Ra\Lift^{\Gamma-(1+i)\,(\U\,i)}) \\
    & \Tm\,\Gamma\,A && := \Tm\,\diamond\,\Bigg(\Pi\,(\Lift^{i-\Gamma}\,\Gamma)\,\bigg(\Lift^{\Gamma-i}\,\Big(\El\, \\
    & && \hspace{4em} \big(\un^{\Gamma-(1+i)}\,(A[\p]\cdot\mk^{1+i-\Gamma}\,(\un^{i-\Gamma}\,\q))\big)\Big)\bigg)\Bigg)
  To make the notation readable, we will not write the lifting
  decorations from now on, and we repeat the previous definitions
  without mentioning universe levels, but all of what follows is
  definable properly with lifting.
    & \Con && := \Ty\,\diamond \\
    & \Sub\,\Delta\,\Gamma && := \Tm\,\diamond\,(\Delta\Ra\Gamma) \\
    & \gamma\circ\delta && := \lam\,(\gamma[\p]\cdot(\delta[\p]\cdot\q)) \\
    & \id && := \lam\,\q \\
    & \diamond && := \top \\
    & \epsilon && := \lam\,\tt \\
    & \Ty\,\Gamma && := \Tm\,\diamond\,(\Gamma\Ra\U) \\
    & A[\gamma] && := \lam\,(A[\p]\cdot(\gamma[\p]\cdot\q)) \\
    & \Tm\,\Gamma\,A && := \Tm\,\diamond\,(\Pi\,\Gamma\,(\El\,(A[\p]\cdot\q))) \\
    & a[\gamma] && := \lam\,(a[\p]\cdot(\gamma[\p]\cdot\q)) \\
    & \Gamma\ext A && := \Sigma\,\Gamma\,(\El\,(A[\p]\cdot\q)) \\
    & (\gamma,a) && := \lam\,(\gamma[\p]\cdot\q,a[\p]\cdot\q) \\
    & \p && := \lam\,(\fst\,\q) \\
    & \q && := \lam\,(\snd\,\q)
  Context extension is given by $\Sigma$ types and pairing/projections
  by pairing/projections of $\Sigma$. All the CwF equations hold, for
  example we derive the functor law for type substitution as follows.
    & A[\gamma\circ\delta] && {=} \\
    & \lam\,(A[\p]\cdot(\lam\,(\gamma[\p]\cdot(\delta[\p]\cdot\q))[\p]\cdot\q)) && {=}(\lam[]) \\
    & \lam\,(A[\p]\cdot(\lam\,(\gamma[\p][\p^+]\cdot(\delta[\p][\p^+]\cdot\q))\cdot\q)) && {=}(\Pi\beta) \\
    & \lam\,(A[\p]\cdot(\gamma[\p]\cdot(\delta[\p]\cdot\q))) && {=}(\Pi\beta) \\
    & \lam\,(\lam\,(A[\p][\p^+]\cdot(\gamma[\p][\p^+]\cdot\q))\cdot(\delta[\p]\cdot\q))\,\, && {=}(\lam[]) \\
    & \lam\,(\lam\,(A[\p]\cdot(\gamma[\p]\cdot\q))[\p]\cdot(\delta[\p]\cdot\q)) && {=} \\
    & A[\gamma][\delta]
  Type formers are added by adjusting them to handle contexts built up
  by $\Sigma$ types, for example,
  \[
  \Pi\,A\,B := \lam\,(\cd\,(\Pi\,(\El\,(A[\p]\cdot\q))\,(\El\,(B[\p][\p]\cdot(\q[\p],\q)))))
  \]
  and $\lam\,b := \lam\,(\lam\,(b[\p][\p]\cdot(\q[\p],\q)))$. All
  substitution laws and $\beta$/$\eta$-laws hold.
\end{proof}
The notion of equivalence between CwFs that we use is contextual
isomorphism \cite{DBLP:journals/corr/abs-2211-07487}: a weak
CwF-morphism (pseudomorphism) which is bijective on types and
terms. Note that a conextual isomorphism preserves all type formers
specified by universal properties.
\begin{problem}
  Given a CwF (with type formers) $M$, let $M'$ denote the CwF
  obtained by first seeing it as an SSC and then termifying it. There
  is a contextual isomorphism between $M'$ and $M$.
\end{problem}
\begin{proof}[Construction]
  We denote the components of the contextual isomorphism $F$ as follows.
    & F : \Con_{M'}\ra\Con_M \\
    & F : \Sub_{M'}\,\Delta\,\Gamma\ra\Sub_M\,(F\,\Delta)\,(F\,\Gamma) \\
    & F : \Ty_{M'}\,\Gamma \cong \Ty_M\,(F\,\Gamma) \\
    & F : \Tm_{M'}\,\Gamma\,A \cong \Tm_M\,(F\,\Gamma)\,(F\,A)
  We define them in the same order as follows, omitting the $M$ subscripts.
    & F : \Ty\,\diamond\ra\Con \\
    & F\,\Gamma := \diamond\ext\Gamma \\
    & F : \Tm\,\diamond\,(\Delta\Ra\Gamma) \ra \Sub\,(\diamond\ext\Delta) (\diamond\ext\Gamma) \\
    & F\,\gamma := (\p,\gamma[\p]\cdot\q) \\
    & F : \Tm\,\diamond\,(\Gamma\Ra\U) \cong \Tm\,(\diamond\ext\Gamma)\,\U \cong \Ty\,(\diamond\ext\Gamma) \\
    & F\,A := \El\,(A[\p]\cdot\q) \\
    & F : \Tm\,\diamond\,(\Pi\,\Gamma\,(\El\,(A[\p]\cdot\q))) \cong \Tm\,(\diamond\ext\Gamma)\,(\El\,(A[\p]\cdot\q)) \\
    & F\,a := a[\p]\cdot\q
  It is easy to check that $F$ preserves the CwF structure, i.e.\ it
  is functor, $\epsilon : \Sub\,(F\,\diamond)\,\diamond$ is an
  isomorphism, $F\,(A[\gamma]) = F\,A[F\,\gamma]$, $F\,(a[\gamma]) =
  F\,a[F\,\gamma]$ and $(F\,\p,F\,\q) : \Sub\,(F\,(\Gamma\ext
  A))\,(F\,\Gamma\ext F\,A)$ is an isomorphism.
\end{proof}
Starting with an SSC-model, termifying it and comparing the result
with the original SSC does not provide an isomorphism because we can't
define the map $F : \Tm\,\diamond\,(\Delta\Ra\Gamma) \ra
\Sub\,(\diamond\ext\Delta)\,(\diamond\ext\Gamma)$ on substitutions is
not definable using single substitutions. We leave formulating the
right notion of contextual isomorphism for SSCs as future work.

@@

If we have var, then we cannot do termification to get a CwF from a var-SSC.

Prove full equivalence of CwF+ with SSC+.

@@

Definition \ref{def:tt} in full detail: \\
The core substitution calculus:
\begin{alignat*}{10}
  & \Con && : \Set \\
  & \Ty && : \Con\ra\N\ra\Set \\
  & \diamond && : \Con \\
  & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con \\
  & \Sub && : \Con\ra\Con\ra\Set \\
  & \Tm && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set \\
  & \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
  & \langle\blank\rangle && : \Tm\,\Gamma\,A\ra\Sub\,\Gamma\,(\Gamma\ext A) \\
  & \blank^+ && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A) \\
  & \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \\
  & \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma]) \\
  & \q && : \Tm\,(\Gamma\ext A)\,(A[\p]) \\
  & [\p][^+] && : B[\p][\gamma^+] = B[\gamma][\p] \\
  & [\p][^+] && : b[\p][\gamma^+] = b[\gamma][\p] \\
  & \q[^+] && : \q[\gamma^+] = \q \\
  & [\p][\langle\rangle] && : B[\p][\langle a\rangle] = B \\
  & [\p][\langle\rangle] && : b[\p ][\langle a\rangle] = b \\
  & \q[\langle\rangle] && : \q[\langle a\rangle] = a \\
  & [\langle\rangle][] && : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle] \\
  & [\p^+][\langle\q\rangle] && : B[\p^+][\langle\q\rangle] = B
\end{alignat*}
Rules for individual type formers.
\begin{alignat*}{10}
  & \Pi && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
  & \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+]) \\
  & \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
  & \lam[] && : (\lam\,b)[\gamma] = \lam\,(b[\gamma^+]) \\
  & \blank\cdot\blank && : \Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,(B[\langle a\rangle]) \\
  & {\cdot}[] && : (t\cdot a)[\gamma] = (t[\gamma])\cdot(a[\gamma]) \\
  & \Pi\beta && : \lam\,b\cdot a = b[\langle a\rangle] \\
  & \Pi\eta && : t = \lam\,(t[\p]\cdot\q) \\
  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i) \\
  & \U[] && : (\U\,i)[\gamma] = \U\,i \\  
  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i \\
  & \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) \\
  & \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma]) \\
  & \U\beta && : \El\,(\cd\,A) = A \\
  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A} \\
  & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i) \\
  & \Lift[] && : (\Lift\,A)[\gamma] = \Lift\,(A[\gamma])
\end{alignat*}
\begin{alignat*}{10}
  & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) \\
  & \mk[] && : (\mk\,a)[\gamma] = \mk\,(a[\gamma]) \\
  & \un[] && : (\un\,a)[\gamma] = \un\,(a[\gamma]) \\
  & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A \\
  & \Lift\beta && : \un\,(\mk\,a) = a \\
  & \Lift\eta && : \mk\,(\un\,a) = a
\end{alignat*}
